---
title: "Hands-on Exercise 7: Global / Local Measures of Spatial Autocorrelation"
date: "19 Feb 2023"
date-modified: "`r Sys.Date()`"
format: html
execute: 
  message: true 
  warning: false
editor: visual
---

# Overview

Objectives

-   import geospatial data using appropriate function(s) of sf package,

-   import csv file using appropriate function of readr package,

-   perform relational join using appropriate join function of dplyr package,

-   compute Global Spatial Autocorrelation (GSA) statistics by using appropriate functions of spdep package,

    -   plot Moran scatterplot,

    -   compute and plot spatial correlogram using appropriate function of spdep package.

-   compute Local Indicator of Spatial Association (LISA) statistics for detecting clusters and outliers by using appropriate functions **spdep** package;

-   compute Getis-Ord's Gi-statistics for detecting hot spot or/and cold spot area by using appropriate functions of **spdep** package; and

-   to visualise the analysis output by using **tmap** package.

# Getting Started

## Analytical Question

In spatial policy, one of the main development objective of the local government and planners is to ensure equal distribution of development in the province. The task in this study is to apply appropriate spatial statistical methods to discover if development are even distributed geographically.

If distribution is uneven, then we should ask "are there signs of spatial clustering?". If distribution is even, then we should ask "where are these clusters?".

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of [Hunan Province](https://en.wikipedia.org/wiki/Hunan), People Republic of China.

## The Data

Two data sets will be used in this hands-on exercise, they are:

-   Hunan province administrative boundary layer at county level. This is a geospatial data set in ESRI shapefile format.

-   Hunan_2012.csv: This csv file contains selected Hunan's local development indicators in 2012.

## R Packages

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

# Import Data

Geospatial data

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

Aspatial

```{r}
hunan2012<- read_csv("data/aspatial/Hunan_2012.csv")
```

Relational Join - joins the geospatial and aspatial data by left join on the field 'County' - retains only the 1st to 4th field, 7th, and 15th field

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

Visualising GDPPC on tmap

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

# Global Spatial Autocorrelation

Spatial Autocorrelation measures the degree of similarity between nearby objects. One of the main reasons why spatial auto-correlation is important is because statistics rely on observations being independent of one another. If autocorrelation exists in a map, then **this violates the fact that observations are independent of one another**.

-   Positive spatial autocorrelation is when similar values cluster together on a map.

-   Negative spatial autocorrelation is when dissimilar values cluster together on a map.

-   A checkerboard pattern is an example where Moran\'s I is -1 because dissimilar values are next to each other.

Spatial autocorrelation measures how close objects are in comparison with other close objects. Moran\'s I can be classified as positive, negative, and with no spatial auto-correlation.

[Useful references on spatial autocorrelation]{.underline}

-   https://rspatial.org/analysis/3-spauto.html

-   https://gisgeography.com/spatial-autocorrelation-moran-i-gis/

-   https://mgimond.github.io/Spatial/spatial-autocorrelation.html

## Computing Contiguity Spatial Weights

Before we can compute the global spatial autocorrelation statistics, we need to construct spatial weights of the study area. The spatial weights are used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In the code chunk below, [poly2nb()](https://r-spatial.github.io/spdep/reference/poly2nb.html) of spdep package is used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. If you look at the documentation you will see that you can pass a "queen" argument that takes TRUE or FALSE as options. If you do not specify this argument the default is set to TRUE, that is, if you don't specify queen = FALSE this function will return a list of first order neighbours using the Queen criteria. If queen=TRUE, a single shared boundary point meets the contiguity condition, if queen=FALSE, more than one shared point is required.

More specifically, the code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbours.

## Row Standardised Weight Matrix

Next, we need to assign weights to each neighboring polygon using the [nb2listw()](https://r-spatial.github.io/spdep/reference/nb2listw.html) of spdep. In our case, each neighboring polygon will be assigned equal weight (style="W"). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values.

While this is the most intuitive way to summaries the neighbors' values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons, thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we'll stick with the style="W" option for simplicity's sake but note that other more robust options are available, notably style="B".

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

The input of nb2listw() must be an object of class nb. The syntax of the function has two major arguments, namely style and zero.poly.

-   style can take values "W", "B", "C", "U", "minmax" and "S". B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   If zero policy is set to TRUE, weights vectors of zero length are inserted for regions without neighbour in the neighbours list. These will in turn generate lag values of zero, equivalent to the sum of products of the zero row t(rep(0, length=length(neighbours))) %\*% x, for arbitrary numerical vector x of length length(neighbours). The spatially lagged value of x for the zero-neighbour region will then be zero, which may (or may not) be a sensible choice.

## Global Spatial Autocorrelation: Moran I

Using the [moran.test()](https://r-spatial.github.io/spdep/reference/moran.test.html) of spdep, Moran I test is performed, which is a statistical test for spatial autocorrelation.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

### Monte Carlo test

Another method is the Monte Carlo Moran I's test using the [moran.mc()](https://r-spatial.github.io/spdep/reference/moran.mc.html) of spdep performed on a total of 1000 simulations.

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

### Visualising Monte Carlo Moran I

It is always a good practice for us the examine the simulated Moran\'s I test statistics in greater detail. This can be achieved by plotting the distribution of the statistical values as a histogram by using the code chunk below.

In the code chunk below [*hist()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/hist) and [*abline()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/abline) of R Graphics are used.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

## Global Spatial Auto Correlation: Geary's

The code chunk below performs Geary\'s C test for spatial autocorrelation by using [*geary.test()*](https://r-spatial.github.io/spdep/reference/geary.test.html) of **spdep**.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

### Monte Carlo Geary's C

Using using [*geary.mc()*](https://r-spatial.github.io/spdep/reference/geary.mc.html) of **spdep**.

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

### Visualising Monte Carlo Geary's C

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

# Spatial Correlogram

Spatial correlograms are great to examine patterns of spatial autocorrelation in your data or model residuals. They show how correlated are pairs of spatial observations when you increase the distance (lag) between them - they are plots of some index of autocorrelation (Moran\'s I or Geary\'s c) against distance.Although correlograms are not as fundamental as variograms (a keystone concept of geostatistics), they are very useful as an exploratory and descriptive tool. For this purpose they actually provide richer information than variograms.

## Compute Moran I's Correlogram

In the code chunk below, [*sp.correlogram()*](https://r-spatial.github.io/spdep/reference/sp.correlogram.html) of **spdep** package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Moran\'s I. The **plot()** of base Graph is then used to plot the output.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

Interpretation ([reference](https://stats.stackexchange.com/questions/182627/interpretation-of-correlogram)):

-   On the y axis is the autocorrelation. The x axis tells you the lag. So, if x=1 we are looking at the correlation of 1st and 2nd, 2nd and 1st counties, etc. If x=2, we have a lag of 2 and we are looking at the correlation of 2nd and 3rd, 3rd with 4th, etc.

-   Positive correlation for the first 4 lags, negative correlation for last 2 lags

By plotting the output might not allow us to provide complete interpretation. This is because not all autocorrelation values are statistically significant. Hence, it is important for us to examine the full analysis report by printing out the analysis results as in the code chunk below.

```{r}
print(MI_corr)
```

## Compute Geary's C Correlogram

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```

# Local Measures of Spatial Autocorrelation

The Local Moran's I statistic is relatively similar to the Global Moran's *I* in that it is providing a measure of how similar locations are to their neighbours. However, the difference is that each location, *i*, receive its own *I* value, as well as its own variance, z value, expected *I*, and variance of *I*.

Reference: https://storymaps.arcgis.com/stories/5b26f25bb81a437b89003423505e2f71

## Cluster & Outlier Analysis

Local Indicators of Spatial Association (LISA) are statistics that **evaluate the existence of clusters** in the spatial arrangement of a given variable. For instance if we are studying cancer rates among census tracts in a given city local clusters in the rates mean that there are areas that have higher or lower rates than is to be expected by chance alone; that is, the values occurring are above or below those of a random distribution in space.

In this section, you will learn how to apply appropriate Local Indicators for Spatial Association (LISA), especially local Moran\'I to detect cluster and/or outlier from GDP per capita 2012 of Hunan Province, PRC.

### Compute Local Moran I

To compute local Moran\'s I, the [*localmoran()*](https://r-spatial.github.io/spdep/reference/localmoran.html) function of **spdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran\'s I of *GDPPC2012* at the county level.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

*localmoran()* function returns a matrix of values whose columns are:

-   Ii: the local Moran\'s I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

The code chunk below list the content of the local Moran matrix derived by using [*printCoefmat()*](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/printCoefmat).

```{r}
printCoefmat(data.frame(
  localMI[fips,], 
  row.names=hunan$County[fips]),
  check.names=FALSE)
```

### Visualising Local Moran I

Before mapping the local Moran\'s I map, it is wise to append the local Moran\'s I dataframe (i.e. localMI) onto hunan SpatialPolygonDataFrame. The code chunks below can be used to perform the task. The out SpatialPolygonDataFrame is called *hunan.localMI*.

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

Mapping the Local Moran I statistic using tmap:

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

Mapping the Local Moran I p values: The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values. The code chunks below produce a choropleth map of Moran\'s I p-values by using functions of **tmap** package.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

Comparing p values and Local Moran I values

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

# Creating LISA Cluster Map

The LISA Cluster Map shows the **significant locations** color coded by **type of spatial autocorrelation**. The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

## Plotting Moran I Scatterplot

The Moran scatterplot is an illustration of the relationship between the values of the chosen attribute at each location and the average value of the same attribute at neighboring locations.

The code chunk below plots the Moran scatterplot of GDPPC 2012 by using [*moran.plot()*](https://r-spatial.github.io/spdep/reference/moran.plot.html) of **spdep**.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

Interpretation: The plot is split in 4 quadrants. The top right corner belongs to areas that have high GDPPC and are surrounded by other areas that have the average level of GDPPC. This are the high-high locations in the lesson slide.

## Plotting Moran Scatterplot With Standardised Variable

First we will use [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% 
  as.vector 
```

The [*as.vector()*](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/as.vector) added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

## Preparing LISA Map Classes

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC) #derive the spatially lagged variable of interest (GDPPC) and center the spatially lagged variable around its mean
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC) 

LM_I <- localMI[,1]  # Do the same for local Moran's around the mean

signif <- 0.05       # Set statistical significance level for local Moran I

# define the low-low (1), low-high (2), high-low (3) and high-high (4) categories
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4    

quadrant[localMI[,5]>signif] <- 0 # place non-significant Moran in the category 0
```

## Plotting LISA Map

```{r}
hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)
```

For effective interpretation, it is better to plot both the local Moran\'s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, 
             asp=1, ncol=2)
```

> show full colour palette spectrum?

# Hot & Cold Spot Area Analysis

Beside detecting cluster and outliers, localised spatial statistics can be also used to detect hot spot and/or cold spot areas.

The term \'hot spot\' has been used generically across disciplines to describe a region or value that is higher relative to its surroundings (Lepers et al 2005, Aben et al 2012, Isobe et al 2015).

## Getis & Ord's G-statistic

An alternative spatial statistics to detect spatial anomalies is the Getis and Ord\'s G-statistics (Getis and Ord, 1972; Ord and Getis, 1995). It looks at neighbours within a defined proximity to identify where either high or low values clutser spatially. Here, statistically significant hot-spots are recognised as areas of high values where other areas within a neighbourhood range also share high values too.

The analysis consists of three steps:

-   Deriving spatial weight matrix

-   Computing Gi statistics

-   Mapping Gi statistics

## Deriving Distance-Based Weight Matrix

First, we need to define a new set of neighbours. Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

-   fixed distance weight matrix; and

-   adaptive distance weight matrix

### Derive Centroid

We will need points to associate with each polygon before we can make our connectivity graph. It will be a little more complicated than just running *st_centroid()* on the sf object: **us.bound**. We need the coordinates in a separate data frame for this to work. To do this we will use a mapping function. The mapping function applies a given function to each element of a vector and returns a vector of the same length. Our input vector will be the geometry column of us.bound. Our function will be *st_centroid()*. We will be using map_dbl variation of map from the purrr package. For more documentation, check out map documentation

To get our longitude values we map the *st_centroid()* function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1. This allows us to get only the longitude, which is the first value in each centroid.

```{r}
longitude <- map_dbl(hunan$geometry,  ~st_centroid(.x)[[1]])
```

```{r}
latitude <- map_dbl(hunan$geometry,  ~st_centroid(.x)[[2]])
```

```{r}
coords <- cbind(longitude, latitude)
```

### Derive Cut-off Distance

Firstly, we need to determine the upper limit for distance band by using the steps below:

-   Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [*knearneigh()*](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.

-   Convert the knn object returned by *knearneigh()* into a neighbours list of class nb with a list of integer vectors containing neighbour region number ids by using [*knn2nb()*](https://r-spatial.github.io/spdep/reference/knn2nb.html).

-   Return the length of neighbour relationship edges by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.

-   Remove the list structure of the returned object by using [**unlist()**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

### Fixed 

Now, we will compute the distance weight matrix by using [*dnearneigh()*](https://r-spatial.github.io/spdep/reference/dnearneigh.html) as shown in the code chunk below.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

Next, convert the nb object to a spatial weights object using nb2listw()

```{r}
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

### Adaptive

One of the characteristics of fixed distance weight matrix is that **more densely settled** areas (usually the urban areas) tend to have **more neighbours** and the **less densely settled** areas (usually the rural counties) tend to have **less neighbours**. Having many neighbours **smoothes** the neighbour relationship across more neighbours.

It is possible to control the number of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
```

```{r}
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

# Computing Gi Statistic

A great introduction to the Gi statistic can be found on the [ArcGIS Pro website](https://pro.arcgis.com/en/pro-app/latest/tool-reference/spatial-statistics/h-how-hot-spot-analysis-getis-ord-gi-spatial-stati.htm):

The [Hot Spot Analysis](https://pro.arcgis.com/en/pro-app/3.1/tool-reference/spatial-statistics/hot-spot-analysis.htm) tool calculates the Getis-Ord Gi\* statistic (pronounced G-i-star) for each feature in a dataset. The resultant [z-scores and p-values](https://pro.arcgis.com/en/pro-app/3.1/tool-reference/spatial-statistics/what-is-a-z-score-what-is-a-p-value.htm) tell you where features with either high or low values cluster spatially. This tool works by looking at each feature within the context of neighboring features. A feature with a high value is interesting but may not be a statistically significant hot spot. To be a statistically significant hot spot, a feature will have a high value and be surrounded by other features with high values as well. The local sum for a feature and its neighbors is compared proportionally to the sum of all features; when the local sum is very different from the expected local sum, and when that difference is too large to be the result of random chance, a statistically significant [z-score](https://pro.arcgis.com/en/pro-app/3.1/tool-reference/spatial-statistics/what-is-a-z-score-what-is-a-p-value.htm) results. When the [FDR correction](https://pro.arcgis.com/en/pro-app/3.1/tool-reference/spatial-statistics/what-is-a-z-score-what-is-a-p-value.htm#ESRI_SECTION1_2C5DFC8106F84F988982CABAEDBF1440) is applied, statistical significance is adjusted to account for multiple testing and spatial dependency.

The Gi\* statistic returned for each feature in the dataset is a z-score. For statistically significant positive z-scores, the larger the z-score is, the more intense the clustering of high values (hot spot). For statistically significant negative z-scores, the smaller the z-score is, the more intense the clustering of low values (cold spot). For more information about determining statistical significance and correcting for multiple testing and spatial dependency, see [What is a z-score? What is a p-value?](https://pro.arcgis.com/en/pro-app/3.1/tool-reference/spatial-statistics/what-is-a-z-score-what-is-a-p-value.htm)

## Fixed Distance

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

The output of localG() is a vector of G or Gstar values, with attributes \"gstari\" set to TRUE or FALSE, \"call\" set to the function call, and class \"localG\".

The Gi statistics is represented as a **Z-score**. Greater values represent a **greater intensity of clustering** and the **direction** (positive or negative) indicates high or low clusters.

Next, we will join the Gi values to their corresponding hunan sf data frame by using the code chunk below.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

In fact, the code chunk above performs three tasks.

-   First, it convert the output vector (i.e *gi.fixed*) into r matrix object by using *as.matrix()*.

-   Next, *cbind()* is used to join hunan\@data and *gi.fixed* matrix to produce a new SpatialPolygonDataFrame called *hunan.gi*.

-   Lastly, the field name of the gi values is renamed to *gstat_fixed* by using *rename()*.

### Mapping Gi Values

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

## Adaptive Distance

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

### Mapping Gi Values

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```
