---
title: "Hands-on Exercise 9: Geographical Segmentation with Spatially Constrained Clustering Techniques"
date: "1 Mar 2023"
date-modified: "`r Sys.Date()`"
format: html
execute: 
  message: true 
  warning: false
editor: visual
---

# Overview

This hands-on exercise contains the following objectives; mainly to introduce the following major analysis types:

1.  Hierarchical cluster analysis
2.  Spatially constrained cluster analysis

With the following learning objectives:

-   to convert GIS polygon data into R\'s simple feature data.frame by using appropriate functions of **sf** package of R;

-   to convert simple feature data.frame into R\'s SpatialPolygonDataFrame object by using appropriate **sf** of package of R;

-   to perform custer analysis by using *hclust()* of Base R;

-   to perform spatially constrained cluster analysis using *skater()* of Base R; and

-   to visualise the analysis output by using **ggplot2** and **tmap** package.

# Getting Started

## Analytical Question

In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data. In this hands-on exercise, we are interested to delineate [Shan State](https://en.wikipedia.org/wiki/Shan_State), [Myanmar](https://en.wikipedia.org/wiki/Myanmar) into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

## The Data

Two data sets will be used in this study. They are:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both data sets are download from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

## R Packages

Before we get started, it is important for us to install the necessary R packages into R and launch these R packages into R environment.

The R packages needed for this exercise are as follows:

-   Spatial data handling

    -   **sf**, **rgdal** and **spdep**

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap**

-   Multivariate data visualisation and analysis

    -   **coorplot**, **ggpubr**, and **heatmaply**

-   Cluster analysis

    -   **cluster**

    -   **ClustGeo**

The code chunks below installs and launches these R packages into R environment.

```{r}
pacman::p_load(rgdal, spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

# Data Import

## Geospatial Data

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

```{r}
shan_sf
```

```{r}
glimpse(shan_sf)
```

## Aspatial Data

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

```{r}
summary(ict)
```

## Derive New Variables

The unit of measurement of the values are number of household. Using these values directly will be biased due to the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

In order to overcome this problem, we will derive the **penetration rate** of each ICT variable by using the code chunk below featuring the **dplyr** package.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

Notice that six new fields have been added into the data.frame. They are RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR.

# Exploratory Data Analysis

## Using Statistical Graphs

We can plot the distribution of the variables (i.e. Number of households with radio) by using appropriate Exploratory Data Analysis (EDA) as shown in the code chunk below.

Histogram is useful to identify the overall distribution of the data values (i.e. left skew, right skew or normal distribution)

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Boxplot is useful for detecting outliers

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Next, we will also plotting the distribution of the newly derived variables (i.e. Radio penetration rate) by using the code chunk below.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

A multi-plot is used to visualise multiple different fields in the data:

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

## Using Choropleth Maps

### Join Geospatial To Aspatial

Before we can prepare the choropleth map, we need to combine both the geospatial data object (i.e. *shan_sf*) and aspatial data.frame object (i.e. *ict_derived*) into one. This will be performed by using the [*left_join*](https://dplyr.tidyverse.org/reference/join.tbl_df.html) function of **dplyr** package. The *shan_sf* simple feature data.frame will be used as the base data object and the *ict_derived* data.frame will be used as the join table.

The code chunks below is used to perform the task. The unique identifier used to join both data objects is *TS_PCODE*.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

## Prepare Choropleth Map

A quick overview of the distribution of radio penetration rate at township level:

```{r}
qtm(shan_sf, "RADIO_PR")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps, one for the **total number** of households (i.e. TT_HOUSEHOLDS.map) and one for the **total number of household with Radio** (RADIO.map) by using the code chunk below.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

Notice that the choropleth maps above clearly show that townships with relatively larger number of households are also showing relatively high numbers of radio ownership.

Now let us plot the choropleth maps showing the dsitribution of total number of households and Radio penetration rate by using the code chunk below.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

Townships with larger number of households have lower penetration rates compared to those with smaller number of households.

# Correlation Analysis

In this section, you will learn how to use [*corrplot.mixed()*](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package to visualise and analyse the correlation of the input variables.

Correlation analysis ensures that there is no dependency or correlation between independent variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.

# Hierarchical Cluster Analysis

## Steps

1.  Extracting the clustering variables from shan_sf geospatial dataset.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Notice that the final clustering variables list does not include variable INTERNET_PR because it is highly correlated with variable COMPUTER_PR (indicated by dark blue on correlation analysis plot with value of 1).

Next, we need to change the rows by township name instead of row number by using the code chunk below

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Notice that the row number has been replaced into the township name.

Now, we will delete the TS.x field by using the code chunk below.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict)
```

2.  Data Standardisation

In general, multiple variables will be used in cluster analysis. It is not unusual their values range are different. In order to avoid the cluster analysis result being baised to clustering variables with large values, it is useful to standardise the input variables before performing cluster analysis.

3.  Min-Max Standardisation

In the code chunk below, *normalize()* of [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/) package is used to stadardisation the clustering variables by using Min-Max method. The *summary()* is then used to display the summary statistics of the standardised clustering variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

Notice that the values range of the Min-max standardised clustering variables are 0-1 now.

4.  Z-score Standardisation

Z-score standardisation can be performed easily by using [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of Base R. The code chunk below will be used to stadardisation the clustering variables by using Z-score method.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

Notice the mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively.

**Note:** [*describe()*](https://www.rdocumentation.org/packages/Hmisc/versions/4.4-0/topics/describe) of [**psych**](https://cran.r-project.org/web/packages/psych/) package is used here instead of *summary()* of Base R because the earlier provides standard deviation.

***Warning: Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.***

## Visualising Standardised Clustering Variables

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

## Computing Proximity Matrix

In R, many packages provide functions to calculate distance matrix. We will compute the proximity matrix by using [*dist()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of R.

*dist()* supports six distance proximity calculations, they are: **euclidean, maximum, manhattan, canberra, binary and minkowski**. The default is *euclidean* proximity matrix.

The code chunk below is used to compute the proximity matrix using *euclidean* method.

```{r}
proxmat <- dist(shan_ict, method="euclidean")
```

```{r}
# | echo: false
proxmat
```

## Computing Hierarchical Clustering

In R, there are several packages provide hierarchical clustering function. In this hands-on exercise, [*hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) of R stats will be used.

*hclust()* employed agglomeration method to compute the cluster. Eight clustering algorithms are supported, they are: ward.D, ward.D2, single, complete, average(UPGMA), mcquitty(WPGMA), median(WPGMC) and centroid(UPGMC).

The code chunk below performs hierarchical cluster analysis using ward.D method. The hierarchical clustering output is stored in an object of class **hclust** which describes the tree produced by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

```{r}
plot(hclust_ward, cex = 0.6)
```

## Selecting Optimal Clustering Algorithm

One of the challenge in performing hierarchical clustering is to identify stronger clustering structures. The issue can be solved by using use [*agnes()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes) function of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like *hclus()*, however, with the *agnes()* function you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

The code chunk below will be used to compute the agglomerative coefficients of all hierarchical clustering algorithms.

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output above, we can see that Ward\'s method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward\'s method will be used.

## Determining Optimal Clusters

Another technical challenge face by data analyst in performing clustering analysis is to determine the optimal clusters to retain.

There are [three](https://statweb.stanford.edu/~gwalther/gap) commonly used methods to determine the optimal clusters, they are:

-   [Elbow Method](https://en.wikipedia.org/wiki/Elbow_method_(clustering))

-   [Average Silhouette Method](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)

-   [Gap Statistic Method](http://www.web.stanford.edu/~hastie/Papers/gap.pdf)

The [**gap statistic**](http://www.web.stanford.edu/~hastie/Papers/gap.pdf) compares the total within **intra-cluster variation** for different values of k with their expected values under null reference distribution of the data. The estimate of the optimal clusters will be value that maximize the gap statistic (i.e., that yields the largest gap statistic). This means that the clustering structure is far away from the random uniform distribution of points.

To compute the gap statistic, [*clusGap()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/clusGap) of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package will be used.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Also note that the [*hcut*](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/) package.

Next, we can visualise the plot by using [*fviz_gap_stat()*](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/) package.

```{r}
fviz_gap_stat(gap_stat)
```

With reference to the gap statistic graph above, the recommended number of cluster to retain is 1. However, it is not logical to retain only one cluster. By examine the gap statistic graph, the **6-cluster** gives the largest gap statistic and should be the next best cluster to pick.

**Note:** In addition to these commonly used approaches, the [NbClust](https://cran.r-project.org/web/packages/NbClust/) package, published by Charrad et al., 2014, provides 30 indices for determining the relevant number of clusters and proposes to users the best clustering scheme from the different results obtained by varying all combinations of number of clusters, distance measures, and clustering methods.

## Interpreting Dendrograms

In the dendrogram displayed above, each leaf corresponds to one observation. As we move up the tree, observations that are similar to each other are combined into branches, which are themselves fused at a higher height.

The height of the fusion, provided on the vertical axis, indicates the (dis)similarity between two observations. The higher the height of the fusion, the less similar the observations are. Note that, conclusions about the proximity of two observations can be drawn only based on the height where branches containing those two observations first are fused. We cannot use the proximity of two observations along the horizontal axis as a criteria of their similarity.

It\'s also possible to draw the dendrogram with a border around the selected clusters by using [*rect.hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/rect.hclust.html) of R stats. The argument *border* is used to specify the border colors for the rectangles.

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

## Visually-driven Hierarchical Clustering 

With [**heatmaply**](https://cran.r-project.org/web/packages/heatmaply/), we are able to build both highly interactive cluster heatmap or static cluster heatmap.

1.  Transform the data frame into matrix format

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

2.  Plot interactive clustering heat map

```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

## Mapping Clusters

With closed examination of the dendragram above, we have decided to retain six clusters.

[*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

Output is called *groups*. It is a *list* object.

In order to visualise the clusters, the *groups* object need to be appended onto *shan_sf* simple feature object.

The code chunk below form the join in three steps:

-   the *groups* list object will be converted into a matrix;

-   *cbind()* is used to append *groups* matrix onto shan_sf to produce an output simple feature object called `shan_sf_cluster`; and

-   *rename* of **dplyr** package is used to rename *as.matrix.groups* field as *CLUSTER*.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

The choropleth map above reveals the clusters are very fragmented. The is one of the major limitation when non-spatial clustering algorithm such as hierarchical cluster analysis method is used.

# Spatially Constrained Clustering: SKATER

In this section, you will learn how to derive spatially constrained cluster by using [*skater()*](https://r-spatial.github.io/spdep/reference/skater.html) method of [**spdep**](https://r-spatial.github.io/spdep/) package.

## Convert SpatialPolygonsDataframe

First, we need to convert `shan_sf` into SpatialPolygonsDataFrame. This is because SKATER function only support **sp** objects such as SpatialPolygonDataFrame.

The code chunk below uses [*as_Spatial()*](https://r-spatial.github.io/sf/reference/coerce-methods.html) of **sf** package to convert *shan_sf* into a SpatialPolygonDataFrame called *shan_sp*.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

## Compute Neighbour List

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We can plot the neighbours list on shan_sp by using the code chunk below. Since we now can plot the community area boundaries as well, we plot this graph on top of the map. The first plot command gives the boundaries. This is followed by the plot of the neighbor list object, with coordinates applied to the original SpatialPolygonDataFrame (Shan state township boundaries) to extract the centroids of the polygons.

These are used as the nodes for the graph representation. We also set the color to blue and specify add=TRUE to plot the network on top of the boundaries.

```{r}
plot(shan_sp, 
     border=grey(.5))
plot(shan.nb, 
     coordinates(shan_sp), 
     col="blue", 
     add=TRUE)
```

Note that if you plot the network first and then the boundaries, some of the areas will be clipped. This is because the plotting area is determined by the characteristics of the first plot. In this example, because the boundary map extends further than the graph, we plot it first.

## Compute Minimum Spanning Tree

1.  Calculating edge costs

Next, [*nbcosts()*](https://r-spatial.github.io/spdep/reference/nbcosts.html) of **spdep** package is used to compute the cost of each edge. It is the distance between it nodes. This function compute this distance using a data.frame with observations vector in each node.

The code chunk below is used to compute the cost of each edge.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

For each observation, this gives the pairwise dissimilarity between its values on the five variables and the values for the neighbouring observation (from the neighbour list). Basically, this is the notion of a generalised weight for a spatial weights matrix.

Next, We will incorporate these costs into a weights object in the same way as we did in the calculation of inverse of distance weights. In other words, we convert the neighbour list to a list weights object by specifying the just computed ***lcosts*** as the weights.

In order to achieve this, [*nb2listw()*](https://r-spatial.github.io/spdep/reference/nb2listw.html) of **spdep** package is used as shown in the code chunk below.

Note that we specify the *style* as **B** to make sure the cost values are not row-standardised.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

2.  Computation

The minimum spanning tree is computed by mean of the [*mstree()*](https://r-spatial.github.io/spdep/reference/mstree.html) of **spdep** package as shown in the code chunk below.

```{r}
shan.mst <- mstree(shan.w)
```

Checking the class and dimension:

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

Note that the dimension is 54 and not 55. This is because the minimum spanning tree consists on n-1 edges (links) in order to traverse all the nodes.

We can display the content of *shan.mst* by using *head()* as shown in the code chunk below.

```{r}
head(shan.mst)
```

The plot method for the MST includes a way to show the observation numbers of the nodes in addition to the edge. As before, we plot this together with the township boundaries. We can see how the initial neighbour list is simplified to just one edge connecting each of the nodes, while passing through all the nodes.

```{r}
plot(shan_sp, border=gray(.5))
plot.mst(shan.mst, 
         coordinates(shan_sp), 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

## Compute Spatially Constrained Clusters

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

The *skater()* takes three mandatory arguments: - the first two columns of the MST matrix (i.e. not the cost), - the data matrix (to update the costs as units are being grouped), and - the number of cuts. Note: It is set to **one less than the number of clusters**. So, the value specified is **not** the number of clusters, but the number of cuts in the graph, one less than the number of clusters.

The result of the *skater()* is an object of class **skater**. We can examine its contents by using the code chunk below.

```{r}
str(clust6)
```

The most interesting component of this list structure is the groups vector containing the labels of the cluster to which each observation belongs (as before, the label itself is arbitary). This is followed by a detailed summary for each of the clusters in the edges.groups list. Sum of squares measures are given as ssto for the total and ssw to show the effect of each of the cuts on the overall criterion.

Check cluster assignment:

```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observations are in each cluster by means of the table command. Parenthetially, we can also find this as the dimension of each vector in the lists contained in edges.groups. For example, the first list has node with dimension 12, which is also the number of observations in the first cluster.

```{r}
table(ccs6)
```

Lastly, plot the pruned tree showing the 5 clusters on top of the township area.

```{r}
plot(shan_sp, border=gray(.5))
plot(clust6, 
     coordinates(shan_sp), 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

## Visualising Clusters On Choropleth Map

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

For easy comparison, it will be better to place both the hierarchical clustering and spatially constrained hierarchical clustering maps next to each other.

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

# Spatially Constrained Clustering: ClustGeo

## About ClustGeo

[**ClustGeo**](https://cran.r-project.org/web/packages/ClustGeo/) package is an R package specially designed to support the need of performing spatially constrained cluster analysis. More specifically, it provides a Ward-like hierarchical clustering algorithm called `hclustgeo()` including spatial/geographical constraints.

In the nutshell, the algorithm uses two dissimilarity matrices D0 and D1 along with a mixing parameter alpha, whereby the value of alpha must be a real number between \[0, 1\]. D0 can be non-Euclidean and the weights of the observations can be non-uniform. It gives the dissimilarities in the **attribute/clustering variable space**. D1, on the other hand, gives the dissimilarities in the **constraint space**. The criterion minimised at each stage is a convex combination of the homogeneity criterion calculated with D0 and the homogeneity criterion calculated with D1.

The idea is then to determine a value of alpha which increases the spatial contiguity without deteriorating too much the quality of the solution based on the variables of interest. This need is supported by a function called `choicealpha()`.

## Ward-like Hierarchical Clustering

ClustGeo package provides function called `hclustgeo()` to perform a typical Ward-like hierarchical clustering just like `hclust()` you learned in previous section.

To perform non-spatially constrained hierarchical clustering, we only need to provide the function a dissimilarity matrix as shown in the code chunk below.

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

Note that the dissimilarity matrix must be an object of class `dist`, i.e. an object obtained with the function `dist()`. For sample code chunk, please refer to [5.7.6 Computing proximity matrix](https://r4gdsa.netlify.app/chap05.html#computing-proximity-matrix).

## Mapping Clusters Formed

Similarly, we can plot the clusters on a categorical area shaded map by using the steps we learned in 5.7.12 Mapping the clusters formed.

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

## Spatially Constrained Hierarchical Clustering

Before we can performed spatially constrained hierarchical clustering, a spatial distance matrix will be derived by using [`st_distance()`](https://r-spatial.github.io/sf/reference/geos_measures.html) of sf package.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

Notice that `as.dist()` is used to convert the data frame into matrix.

Next, `choicealpha()` will be used to determine a suitable value for the mixing parameter alpha as shown in the code chunk below.

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

With reference to the graphs above, alpha = 0.3 will be used as shown in the code chunk below.

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.3)
```

Use cutree() to derive cluster object.

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

Then join back the group list with *shan_sf* polygon feature data frame by using the code chunk below.

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

Plot the newly delineated spatially constrained clusters.

```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```

# Visual Interpretation of Clusters

## Individual Clustering Variable

Code chunk below is used to reveal the distribution of a clustering variable (i.e RADIO_PR) by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

The boxplot reveals Cluster 3 displays the highest mean Radio Ownership Per Thousand Household. This is followed by Cluster 2, 1, 4, 6 and 5.

## Multivariate Visualisation

Past studies shown that parallel coordinate plot can be used to reveal clustering variables by cluster very effectively. In the code chunk below, [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://ggobi.github.io/ggally/) package.

```{r}
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

The parallel coordinate plot above reveals that households in Cluster 4 townships tend to own the highest number of TV and mobile-phone. On the other hand, households in Cluster 5 tends to own the lowest of all the five ICT.

Note that the `scale` argument of `ggparcoor()` provide several methods to scale the clustering variables. They are:

-   std: univariately, subtract mean and divide by standard deviation.

-   robust: univariately, subtract median and divide by median absolute deviation.

-   uniminmax: univariately, scale so the minimum of the variable is zero, and the maximum is one.

-   globalminmax: no scaling is done; the range of the graphs is defined by the global minimum and the global maximum.

-   center: use uniminmax to standardize vertical height, then center each variable at a value specified by the scaleSummary param.

-   centerObs: use uniminmax to standardize vertical height, then center each variable at the value of the observation specified by the centerObsID param

There is no one best scaling method to use. You should explore them and select the one that best meet your analysis need.

Last but not least, we can also compute the summary statistics such as mean, median, sd, etc to complement the visual interpretation.

In the code chunk below, `group_by()` and `summarise()` of dplyr are used to derive mean values of the clustering variables.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```
