---
title: "Hands-on Exercise 11: Modelling Geographical Accessibility"
date: "20 Mar 2023"
date-modified: "`r Sys.Date()`"
format: html
execute: 
  message: true 
  warning: false
editor: visual
---

# Getting Started

New package: </a href="https://cran.r-project.org/web/packages/SpatialAcc/SpatialAcc.pdf"> SpatialAcc <a>

```{r}
pacman::p_load(tmap, SpatialAcc, sf, sp,
               ggstatsplot, reshape2,
               tidyverse)
```

# The Data

Four data sets will be used in this hands-on exercise, they are:

MP14_SUBZONE_NO_SEA_PL: URA Master Plan 2014 subzone boundary GIS data. This data set is downloaded from data.gov.sg.

hexagons: A 250m radius hexagons GIS data. This data set was created by using st_make_grid() of sf package. It is in ESRI shapefile format.

ELDERCARE: GIS data showing location of eldercare service. <a href="https://data.gov.sg/dataset/eldercare-services"> This data</a> is downloaded from data.gov.sg. There are two versions. One in ESRI shapefile format. The other one in Google kml file format. For the purpose of this hands-on exercise, ESRI shapefile format is provided.

OD_Matrix: a distance matrix in csv format. There are six fields in the data file. They are:

-   origin_id: the unique id values of the origin (i.e. fid of hexagon data set.),

-   destination_id: the unique id values of the destination (i.e. fid of ELDERCARE data set.),

-   entry_cost: the perpendicular distance between the origins and the nearest road),

-   network_cost: the actual network distance from the origin and destination,

-   exit_cost: the perpendicular distance between the destination and the nearest road),

-   and total_cost: the summation of entry_cost, network_cost and exit_cost.

::: call-out
The OD matrix will take a long time to run if using road network.
:::

All the values of the cost related fields are in metres.

Reminder: Except MP14_SUBZONE_NO_SEA_PL data set, the other three data set are specially prepared by Prof. Kam for teaching and research purpose. Students taking IS415 Geospatial Analytics and Applications are allowed to use them for hands-on exercise purpose. Please obtain formal approval from Prof. Kam if you want to use them for other courses or usage.

# Geospatial Data Wrangling

Importing geospatial masterplan subzone layer from 2014

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

Importing in the hexagon layer

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

Import the eldercare services layer

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

## Update CRS Information

```{r}
mpsz <- st_transform(mpsz, 3414)
hezagons <- st_transform(hexagons, 3414)
eldercare <- st_transform(eldercare, 3414)
```

```{r}
st_crs(mpsz)
```

# Aspatial Data Wrangling

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

The rows represent origins (i.e. also know as from field) and the columns represent destination (i.e. also known as to field.)

The code chunk below uses spread() of tidyr package is used to transform the O-D matrix from a thin format into a fat format.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

Note: Since tidyr version 1.0 a new function called pivot_wider() is introduce. You should use pivot_wider() instead of spread()

Currently, the distance is measured in metre because SVY21 projected coordinate system is used. The code chunk below will be used to convert the unit f measurement from metre to kilometre.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

# Modelling Accessibility: Hansen Method

[Hansen Accessibility Model](http://frontdesk.co.in/planning/land-use-and-transport-planning/hansen-accessibility-model/#:~:text=Hansen%20Accessibility%20Model%20(1959)%20is,other%20activity%20in%20the%20region%20.) (1959) is based upon concept that the more accessible an area is to various activities and the more vacant land area has greater growth potential. This accessibility of an area is an index representing the closeness of area to all other activity in the region. HAM calculates the accessibility of each zone by considering the distance to other zones and the population or employment base of each zone. The model then generates a series of accessibility values that reflect the ease with which people can reach different parts of the city using different transportation modes.

## Compute Hansen Accessibility

Now, we ready to compute Hansen\'s accessibility by using *ac()* of [**SpatialAcc**](https://cran.r-project.org/web/packages/SpatialAcc/) package. Before getting started, you are encourage to read the [arguments](https://cran.r-project.org/web/packages/SpatialAcc/SpatialAcc.pdf) of the function at least once in order to ensure that the required inputs are available.

The code chunk below calculates Hansen\'s accessibility using *ac()* of **SpatialAcc** and *data.frame()* is used to save the output in a data frame called `acc_Handsen`.

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
```

```{r}

```

```{r}

```

```{r}

```

# Modelling Accessibility: KD2SFCA Method

```{r}

```

```{r}

```

# Modelling Accessibility: Spatial Accessibility Measure Method

```{r}

```
